Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-05-21 paper/manuscript.tex
Water Sampling

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[9]{David F. Hahn}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[6]{Samarjeet Prasad}
\author[2,7]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}
%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}

@@

\paragraph{Conserved binding site waters can play an important role in binding free energies}
Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales. This happens typically in buried binding sites. Overlaying multiple protein X-ray structures can identify conserved or additional water molecules that can be useful to include in calculations. In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}. Other tools such as WaterMap or open source equivalents (SSTMap, GIST, and others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}. Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.

\paragraph{Membrane proteins require additional care}
Integral membrane proteins require embedding in a suitable model membrane to remain stable during simulations. Lipids are usually not included in protein force fields and, although they may be covered by generic small molecule force fields, have often been parameterized separately (e.g., AMBER's Lipid21 \cite{AMBERLipid21} and the CHARMM lipid force field \cite{CHARMMLipidFF}). Membrane systems can be simulated in multiple thermodynamic ensembles, but the semi-isotropic scheme $NP_{z}P_{xy}T$ is frequently employed, allowing the pressure in the bilayer plane ($xy$) to scale independently \cite{LipidBestPracticesCOMS}. These systems require more extensive equilibration than soluble proteins. Some packages, such as Desmond, include a 7-step protocol to relax membrane systems.

Alchemical transformations that involve net charge changes can provide difficulties for membrane systems; phospholipids carry functional groups with nonzero formal charge and cannot be treated as a solvent with a uniform dielectric in the same way as water. Several approaches are reviewed in \cite{wu2022correction-51c} and the recommended method is to use a co-alchemical ion to maintain system neutrality.

In the case of a lipid-facing binding site, there are two additional challenges. First, simulation time should be extended to ensure sufficient sampling of lipid configurational space. Lipids may either bind alongside small molecules, acting as cofactors, or be displaced from the binding site by a ligand. If lipids participate in binding through protein-lipid or ligand-lipid interactions, it is necessary to sample these intermolecular interactions through additional $\lambda$ windows, longer simulation times, or different alchemical protocols. Displacing a lipid will have its own free energy cost which may take a long time to converge if it requires diffusion and rearrangement of nearby lipids. Second, the free energy of binding for a small molecule may be strongly coupled to the lipid species. In some cases, the identity of the lipid may be known. But in other cases, it may be difficult to discern the identity of the lipid from X-ray or cryo-EM structures due to the flexible nature of lipid head groups and benchmarking is recommended to match experimental data.

@@

>>>>

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[9]{David F. Hahn}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[6]{Samarjeet Prasad}
\author[2,7]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}
\author[11]{Oliver J. Melling}
\author[11]{William G. Poole}
\author[11]{Jonathan W. Essex}

%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{School of Chemistry, University of Southampton, Southampton, SO17 1BJ}

@@

\paragraph{Conserved binding site waters can play an important role in binding free energies}
Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales. This can happen in buried binding sites~\cite{laage2017water}. Overlaying multiple protein X-ray structures can identify conserved water molecules or those that are associated only with certain ligands in a series and are therefore important to include in calculations. In cases where water molecules are known to play an important role in ligand binding, for example if the presence of electron density suggests they are tightly bound (indicated by a high EDIA score~\cite{nittinger2015evidence,meyder2017estimating}), software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}. Other tools such as WaterMap~\cite{abel2008role,young2007motifs} or open source equivalents (SSTMap~\cite{haider2018solvation}, GIST~\cite{ramsey2016solvation}, and others) can be used to define the hydration states for systems with no experimental evidence of water sites~\cite{wang2011ligand}. Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.

In the context of relative binding free energy calculations, determining the correct hydration pattern for the two endpoints may not be sufficient for an accurate and reproducible calculation. If the locations of the bound water molecules differs at the two endpoints - i.e. if a perturbation to the ligand displaces a bound water molecule - then the network needs to be able to adapt during the perturbation. Otherwise, it is likely that the estimated relative binding free energies will be dependent on the direction in which the perturbation is performed~\cite{ross2020enhancing}.

In absolute binding free energy calculations, a ligand is fully decoupled from a binding site cavity which may otherwise be hydrated in the \textit{apo} state. In this case, the sampling of water molecules must be sufficient to hydrate the binding pocket as the ligand is decoupled (or vice versa for the reverse). When the binding site is exposed and water can move freely between the site and bulk solvent, then sampling is often not an issue. However, in the case of more buried sites or sites which form complex water networks, enhanced water sampling may be required to ensure an accurate binding affinity.

@@

\subsubsection{Handling poorly sampled water networks}
The location of any water molecules within the binding site of interest must be determined prior to carrying out a free energy calculation. Experimental structures, such as those determined by X-ray and cryo-EM, are often sufficient for this, provided the resolution is low enough. However, we recommend using grand canonical sampling methods\cite{melling2023enhanced, samways2020grand} as a means of verifying the experimental water placement and determining any potential differences owing to the classical forcefield used. Many other water placement methods are available, all varying in their methodology, computational cost and accuracy. Details of these methods can be found in the publication from Samways et al.\cite{samways2021water}.

In relative calculations, if the hydration patterns differ between the two ligand endpoints then careful consideration will need to be given to the adaptation of the water network during the free energy calculation. We recommend performing the free energy calculation in the conventional way, i.e. with no enhanced water sampling, to determine whether the water network can adapt to the ligand perturbation itself. This is likely to happen if the hydration site is not particularly occluded from bulk solvent and water molecules can move freely from bulk into the binding site.

If a conventional sampling method, such as molecular dynamics or Monte Carlo, is unable to rearrange the water network with sufficient efficacy to accommodate the perturbation then an enhanced sampling method will be required to achieve accurate and converged results. One such approach is to include an absolute free energy calculation of the water molecule into the thermodynamic cycle\cite{ross2020enhancing, hamelberg2004standard, barillari2007classification, yu2008free, ge2022abfewat} such that the free energies of both the ligand transformation, and the decoupling of any displaced water molecules, are calculated separately and combined to achieve the total free energy change for the process. This method can be difficult to use in practice as the water molecule needs to be carefully restrained while being decoupled, and care needs to be taken to ensure water does not diffuse back into the hydration site.

Monte Carlo approaches such as the hybrid MC/MD approach proposed by Bergazin et al.\cite{bergazin2021enhancing, ben-shalom2021fast, ge2022enhancing} allow waters to move between bulk solvent and the binding site with greater frequency by proposing MC moves that translate the water molecules within a spherical region that encompasses both bulk solvent and the binding site. This avoids the difficulty of performing a decoupling calculation on the bound waters but can be time-consuming and risk computational time being wasted on move proposals that are not of interest.

Grand canonical (GC) methods propose the addition and removal of water molecules to and from a binding site in accordance with a defined chemical potential. Several implementations of GC moves being coupled to RBFE calculations exist, such as in Schrodinger's FEP+ package, introduced by Ross et al.\cite{ross2020enhancing} with moves proposed both within the binding site (to ensure the correct hydration pattern) and also within bulk solvent (to enhance the sampling of the bulk density). A cavity-bias technique was used for the binding site moves. Grand canonical alchemical perturbation (GCAP)\cite{brucemacdonald2018ligand} is also available in ProtoMS\cite{woods2018protoms}, whereby 2-dimensional titrations are performed in both lambda and chemical potential space. However, given ProtoMS only allows MC sampling to be performed, the configurational sampling of the protein is restricted.

More recently, an OpenMM-based implementation allows for GC moves to be carried out during both equilibrium and nonequilibrium RBFE calculations\cite{melling2025developing}. These moves are performed in the form of grand canonical nonequilibrium candidate Monte Carlo (GCNCMC) moves\cite{melling2023enhanced} and are run during the equilibration phase at each lambda window in the case of equilibrium FEP calculations, or during the nonequilibrium switch in the NES calculations.

Although often water sampling is less of an issue in ABFE calculations, handling changes to water networks in buried pockets during ligand coupling/decoupling can be handled in a similar fashion to RBFEs with grand canonical methods.

% in ABFE calculations is less established, although in principle, the same rules apply. When conventional sampling of water is insufficient such that water does not efficiently bind during ligand decoupling, performing extra ABFE calculations for the removal or addition of water molecules to the thermodynamic cycle is one way of ensuring the contribution of these solvent effects is captured. Performing grand canonical water moves along side a more traditional ABFE calculation of a ligand will ensure that any changes to binding pocket hydration states are sampled \textit{in situ}.

Employing the techniques described above with care should allow the water network to adapt during the perturbation of the ligand to ensure more accurate and converged affinity estimations that are independent of the starting configuration and direction in which the calculation is performed.
